Supplementary materials for 'Canaries of the Arctic: the collapse of eastern Bering Sea snow crab

### 2023-04-16 manuscript_supmat.Rmd
revision commits

---
author: ""
date:
output:
  pdf_document:
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    keep_tex: true
    includes:
      in_header: my_header.tex
header-includes: 
  - \usepackage[left]{lineno}
  - \linenumbers
#-  \pagenumbering{gobble}
csl: fish-and-fisheries.csl
toc: yes
title: "Supplementary materials for 'Canaries of the Arctic: the collapse of eastern Bering Sea snow crab'"
---

@@

# Supplementary materials
## Methods overview
We used an integrated population model to estimate variation in mortality over time for snow crab in the eastern Bering Sea and generalized additive models (GAMs) to relate the estimated variation in mortality to potential stressors in the environment.  The population dynamics model was fit to abundance and size composition data from the National Marine Fisheries Service (NMFS) summer bottom trawl survey on the eastern Bering Sea shelf to estimate total mortality by maturity state and year for male snow crab. We then developed indices for temperature occupied, disease prevalence, cannibalism, and crab density from the NMFS survey to test as covariates in GAMs. Cod predation indices were developed using stomach content data collected on NMFS surveys in addition to cod size composition and abundances. Indices for fishery related effects were collated from fisheries statistics from the Alaska Department of Fish and Game and also included in the GAMs.

@@

### Does incorporating the uncertainty in the estimates of mortality change model outcomes?
The models presented in the main text use the maximum likelihood estimates of mortality from the population dynamics models as response variables in the GAMs. However, each of those estimates of mortality have associated uncertainty estimated in the fitting process. One way of evaluating the impact of incorporating the estimated uncertainty from the population dynamics model into the GAM fitting process can be accomplished in 4 steps:

@@

The model in the main text is unconstrained with respect to the shape of the estimated relationships between mortality and covariates. However, the relationship between immature mortality and mature population size was markedly dome-shaped and a satisfying biological explanation for this shape is not immediately apparent. To explore the impacts of unconstrained estimation of this relationship, we refit the models using shape constrained additive models in the R package 'scam' (based on Pya and Wood, 2015). This allows the user to specify constraints on the shape of relationships between model variables (e.g. monotonically increasing or decreasing). We refit our model for immature mortality with the assumption that the relationship between immature mortality and mature population size can only be monotonically increasing. The rest of the covariates were specified as linear predictors, except temperature, which remained non-linear. Given the importance of temperature in the hypotheses we present, we were particularly interested to understand how the assumptions about the shape of other significant covariates influence the estimated relationship between temperature and mortality.

@@

\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.0325 & 0.1850 & -0.1757 & 0.8623 \\ 
  disease & 0.0236 & 0.0162 & 1.4561 & 0.1607 \\ 
  imm\_pop & -0.0374 & 0.0174 & -2.1454 & 0.0442 \\ 
  predation & 0.0002 & 0.0169 & 0.0137 & 0.9892 \\ 
  bycatch & -0.0003 & 0.0147 & -0.0194 & 0.9847 \\ 
  cannibalism & -0.0090 & 0.0153 & -0.5887 & 0.5626 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(temperature) & 1.0005 & 1.0009 & 6.7766 & 0.0167 \\ 
  s(mat\_pop) & 1.7664 & 2.0738 & 3.5320 & 0.0466 \\ 
   \hline
\end{tabular}
\caption{GAM output for trimmed model predicting immature mortality. Deviance explained =  59.53 \%} 
\label{tab.gam5}
\end{table}

@@

## A word on methods
Attribution of changes in population processes in ecology is a difficult problem, particularly for wild populations that are difficult to directly observe and impossible to experiment on in situ. There are a large range of methodologies that claim to identify causality in observational data (structural equation modeling, empirical dynamic modeling, etc.). Some of the difficulties in determining causality in ecological time series are related to the generally short time series that are available, non-linear dynamics, and departures of populations or covariates into unexplored parameter space. These issues can present issues for any modeling framework and we have tried to address these the the best of our ability with the models used here. The use of p-values has been (rightfully) criticized in the literature and the explanatory power of our models are likely overstated.  Ultimately, the numerous sensitivities and simulation tests performed here were undertaken to try to understand if a suite of covariates appear to be important under different modeling decisions and considerations of uncertainty, in spite of the potential short-comings of the data available and models selected. Temperature and population density proved to be these covariates.

@@

### Are you sure the collapse wasn't a result of trawling?
The bycatch index steadily declined since the beginning of our study, with the relative impact of trawling in 2018 and 2019 below the historical average (\autoref{grid_covar}). It is difficult to reconcile the idea that trawling could have contributed to the collapse with the relatively low mortality rates estimated during the periods when the bycatch index was many times higher in the 1990s. Furthermore, if trawling is a large source of mortality for snow crab, it is difficult to understand how the largest pseudocohort ever observed could have established and survived for ~8 years on the Bering Sea shelf, during which the trawling pressure was relatively consistent.

@@

### What do crab eat? If they starved, did there appear to be large declines in their prey base?
Snow crab have a wide-ranging diet of bivalves, polychaetes, crustaceans, and gastropods in the northern Bering Sea (Kolts et al., 2013). They appear to be a generalist, consuming whatever they can capture and crush with their claws. Kolts et al. (2013) reported that most prey items were consumed in proportion to their estimated abundances, except polychaetes, which seemed to be preferentially selected. The prey items are relatively poorly sampled in the Bering Sea survey, so time series of prey quantity are unreliable.

Even if there were abundant forage for snow crab in warm years, the metabolic trade-off between the energy required to obtain, handle, and digest their prey and the energy derived from prey would need to be considered when trying to understand if metabolic demands could be met. This would be a useful area of further research, particularly if reliable time series of benthic forage could be started and maintained.

@@

### Were that many crab really in the eastern Bering Sea to begin with? Was the 'collapse' an artifact of some survey error?
The NMFS summer survey was designed to estimate crab abundances (REFERENCE). Snow crab are widely distributed on the shelf and consequently well sampled by the survey. There are 375 survey locations in the NMFS eastern Bering Sea trawl survey, of which 349 are on a 40 square nautical mile grid. The remaining stations are in high density sampling areas around islands in the Bering Sea implemented to better estimate crab abundances around those islands. On average, snow crab are observed at 233 of the 349 survey stations on the standard grid. In 2018 a large number of stations returned estimated high densities of crab (see Fig. 1 of main text), which means that the large estimates of abundance in 2018 were not driven by one or two large survey tows. Further, the recent survey methodology has been repeatedly verified as a useful tool for estimating crab abundance (see Somerton et al., 2013, for example).

@@

```{r,echo=FALSE,warning=FALSE,message=F,out.width="98%",fig.cap="\\label{slope}Map of eastern Bering Sea slope habitat (colors). Reproduced from HERE. "}

include_graphics("plots/ebs_slope.png")

```{r,echo=FALSE,warning=FALSE,message=F,out.width="98%",fig.cap="\\label{fish_cpue}Fishery cpue (top; black lines are median, grey box represents 25-75th quantiles, circles are outliers) and number of crab caught in the directed snow crab fishery (bottom). Vertical dashed line represents the introduction of individual transferrable quota management."}

@@

>>>>

---
author: ""
date:
output:
  pdf_document:
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    keep_tex: true
    includes:
      in_header: my_header.tex
header-includes: 
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{float} \floatplacement{figure}{H} 
  - \newcommand{\beginsupplement}{\setcounter{table}{0} \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}

#-  \pagenumbering{gobble}
csl: fish-and-fisheries.csl
toc: yes
title: "Supplementary materials for 'Canaries of the Arctic: the collapse of eastern Bering Sea snow crab'"
---

@@

\beginsupplement
# Supplementary materials
## Methods overview
We used an integrated population model to estimate variation in mortality over time for snow crab in the eastern Bering Sea and generalized additive models (GAMs) to relate the estimated variation in mortality to potential stressors in the environment.  The population dynamics model was fit to abundance and size composition data from the National Marine Fisheries Service (NMFS) summer bottom trawl survey on the eastern Bering Sea shelf to estimate total mortality by maturity state and year for male snow crab. We then developed indices for temperature occupied, disease prevalence, cannibalism, and crab density from the NMFS survey to test as covariates in GAMs. Cod predation indices were developed using stomach content data collected on NMFS surveys in addition to cod size composition and abundances. Indices for fishery related effects were collated from fisheries statistics from the Alaska Department of Fish and Game and also included in the GAMs.

@@

### Does incorporating the uncertainty in the estimates of mortality change analysis outcomes?
The models presented in the main text use the maximum likelihood estimates of mortality from the population dynamics models as response variables in the GAMs. However, each of those estimates of mortality have associated uncertainty estimated in the fitting process. One way of evaluating the impact of incorporating the estimated uncertainty from the population dynamics model into the GAM fitting process can be accomplished in 4 steps:

@@

The model in the main text is unconstrained with respect to the shape of the estimated relationships between mortality and covariates. However, the relationship between immature mortality and mature population size was markedly dome-shaped and a satisfying biological explanation for this shape is not immediately apparent. To explore the impacts of unconstrained estimation of this relationship, we refit the models using shape constrained additive models in the R package 'scam' (based on Pya and Wood, 2015). This allows the user to specify constraints on the shape of relationships between model variables (e.g. monotonically increasing or decreasing). We refit our model for immature mortality with the assumption that the relationship between immature mortality and mature population size can only be monotonically increasing (similar to the estimated relationship between mature mortality and mature population size). The rest of the covariates were specified as linear predictors, except temperature, which remained a non-linear smooth. Given the importance of temperature in the hypotheses we present, we were particularly interested to understand how the assumptions about the shape of other significant covariates influence the estimated relationship between temperature and mortality.

@@

\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.0325 & 0.1850 & -0.1757 & 0.8623 \\ 
  disease & 0.0236 & 0.0162 & 1.4561 & 0.1607 \\ 
  imm\_pop & -0.0374 & 0.0174 & -2.1454 & 0.0442 \\ 
  predation & 0.0002 & 0.0169 & 0.0137 & 0.9892 \\ 
  bycatch & -0.0003 & 0.0147 & -0.0194 & 0.9847 \\ 
  cannibalism & -0.0090 & 0.0153 & -0.5887 & 0.5626 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(temperature) & 1.0005 & 1.0009 & 6.7766 & 0.0167 \\ 
  s(mat\_pop) & 1.7664 & 2.0738 & 3.5320 & 0.0466 \\ 
   \hline
\end{tabular}
\caption{GAM output for a shape constrained additive model predicting immature mortality. Deviance explained =  57.5 \%} 
\label{tab.gam5}
\end{table}

@@

## A word on methods
Attribution of changes in population processes in ecology is a difficult problem, particularly for wild populations that are difficult to directly observe and impossible to experiment on in situ. There are a large range of methodologies that claim to identify causality in observational data (structural equation modeling, empirical dynamic modeling, etc.). Some of the difficulties in determining causality in ecological time series are related to the generally short time series that are available, non-linear dynamics, and departures of populations or covariates into unexplored parameter space. These issues can present issues for any modeling framework and we have tried to address these the the best of our ability with the models used here. The use of p-values has been (rightfully) criticized in the literature and the explanatory power of our models are likely overstated given these criticisms.  Ultimately, the numerous sensitivities and simulation tests performed here were undertaken to try to understand if a suite of covariates appear to be important under different modeling decisions and considerations of uncertainty, in spite of the potential short-comings of the data available and models selected. Temperature and population density appear to be these covariates.

@@

### Are you sure the collapse wasn't a result of trawling?
The bycatch index steadily declined since the beginning of our study, with the relative impact of trawling in 2018 and 2019 below the historical average (\autoref{grid_covar}). It is difficult to reconcile the idea that trawling could have contributed to the collapse with the relatively low mortality rates estimated during the periods when the bycatch index was many times higher in the 1990s. Furthermore, if trawling were a large source of mortality for snow crab, it is difficult to understand how the largest pseudocohort ever observed could have established and survived for ~8 years on the Bering Sea shelf, during which the trawling pressure was relatively consistent.

@@

### What do crab eat? If they starved, did there appear to be large declines in their prey base?
Snow crab have a wide-ranging diet of bivalves, polychaetes, crustaceans, and gastropods in the northern Bering Sea (Kolts et al., 2013). They appear to be a generalist, consuming whatever they can capture and crush with their claws. Kolts et al. (2013) reported that most prey items were consumed in proportion to their estimated abundances, except polychaetes, which seemed to be preferentially selected. Indices of prey biomass can be calculated from the NMFS trawl, however the survey gear does not sample these species well. With that caveat, the average of three of the clusters of scaled prey biomass by group from the survey were below the long-term mean in 2018 and 2019 (\autoref{benthic_inverts}). While this is suggestive, these time series require further validation before placing any confidence in them and were therefore not used in the main analysis.

Even if these time series were reliable and there appeared to be abundant forage for snow crab in warm years, the metabolic trade-off between the energy required to obtain, handle, and digest their prey and the energy derived from prey would need to be considered when trying to understand if metabolic demands could be met. This would be a useful area of further research, particularly if reliable time series of benthic forage could be identified and maintained.

@@

### Were that many crab really in the eastern Bering Sea to begin with? Was the 'collapse' an artifact of some survey error?
The NMFS summer survey was originally designed to estimate crab abundances (Zimmermann et al., 2009) and the recent survey methodology has been repeatedly validated as a useful tool for estimating crab abundance (see Somerton et al., 2013, for example). Snow crab are widely distributed on the shelf and consequently well sampled by the survey. There are 375 survey locations in the NMFS eastern Bering Sea trawl survey, of which 349 are on a 400 square nautical mile grid. The remaining stations are in high density sampling areas around islands in the Bering Sea implemented to better estimate crab abundances around those islands. On average, snow crab are observed at 233 of the 349 survey stations on the standard grid. In 2018 a large number of stations returned estimated high densities of crab (see Fig. 1 of main text), which means that the large estimates of abundance in 2018 were not driven by one or two large survey tows.

@@

```{r,echo=FALSE,warning=FALSE,message=F,out.width="98%",fig.cap="\\label{fish_cpue}Fishery cpue (top; black lines are median, grey box represents 25-75th quantiles, circles are outliers) and number of crab caught in the directed snow crab fishery (bottom). Vertical dashed line represents the introduction of individual transferrable quota management. CPUEs before and after are difficult to compare as a result of changes in fleet behavior."}

@@

```{r,echo=FALSE,warning=FALSE,message=F,out.width="98%",fig.cap="\\label{benthic_inverts}Hierarchical clusters of scaled benthic prey group biomass in the eastern Bering Sea. Blue lines are the fits from a LOESS smoother with a span set to 0.12. Grey lines in the background are the individual scaled time series of observed prey biomass. Names of the species group included in each cluster are to the right of each figure."}

include_graphics("plots/benthic_inverts.png")
